---
title: "chile-aberration"
author: "Dan Weinberger"
date: "February 4, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(surveillance)
library(lubridate)
```

##Goal for the exercise

Today, we will set up an aberration detection algorithm for diarrheal disease among children <5y in Chile (population 18 Million). We will first format and explore the data and will then  test out a few different algorithms and decide on the most appropriate. Administrative hospitalization data for Chile are publicly available on the Ministry of Health Website. Variables included in this parsed down database are:

EDAD: age in years
diag1: primary diagnosis, in ICD10 coding
date: Date admitted to hospital 

You have been provided with a subset of the larger database that includes ICD10 codes that start with the letter "A" (certain infectious diseases). We can learn more about the ICD10 codes here: https://www.icd10data.com/ICD10CM/Codes/A00-B99

### First read in the data 
```{r readscsv}
d1<-read.csv('./aberration detection/chile_under5y_aCHAPTER.csv')
```

## Tell R that 'date' is a date variable and assign a format
see https://www.statmethods.net/input/dates.html
```{r date_format}
d1$date<- as.Date(d1$date, "%d%b%Y")
```

Then create a new variablecalled 'week.date' that has the date of the Sunday of the week in which the admission occurred. The lubridate has some functions that make this easier. You want to round the date down to the nearest Sunday. The floor_data function can accomplish this: https://rawgit.com/rstudio/cheatsheets/master/lubridate.pdf

```{r date_format2}
#"round" the date down to 
d1$week.date<-floor_date(d1$date, unit='week')
```


### Then do some basic explorations. What is the distibution of ages? of Dates? (make a histogram for each)
```{r hist1}
hist(d1$EDAD)
hist(d1$date, breaks=10)
```

### Which codes are the most commonly used in this databse?
Make a table of the codes, sorted most to least common.

```{r freq.codes, echo=FALSE}
sort(table(d1$diag1),decreasing=T)
```


```{r}
icd10.3digits<-substr(d1$diag1,1,3) #extract 1st 3 digits from IC10 code
icd10.3digits[1:10] #view first 10

#Initialize variables
d1$a00_a09<-rep(0, nrow(d1))
d1$a37<-rep(0, nrow(d1))

d1$a00_a09[icd10.3digits %in% c('A00', 'A01', 'A02', 'A03', 'A04', 'A05', 'A06', 'A07', 'A08', 'A09') ]<-1
d1$a37[icd10.3digits %in% c('A37') ]<-1

d1$year<-year(d1$date)
table(d1$year, d1$a00_a09)
table(d1$year, d1$a37)

```

## Let's aggregate now by week.date

```{r}
d1.split<-split(d1, d1$week.date)
d2<- sapply(d1.split, function(x) sum(x$a00_a09) )
```

It is also important to make sure that the time series is 'filled'--if there are weeks with 0 counts, it needs to be represented in the time series. to do this, let's make a vector of week.dates and see if it is of the length as our data
```{r}
dates<-as.Date(sort(names(d2))) #Etract and sort the weekdates
date.seq <- seq.Date(from=dates[1], to=dates[length(dates)], by='week') #create a sequence from first to last week date
length(dates)==length(date.seq) #test if the time series is complete

#data frame 1 has just the time series with all weeks represented 
date.seq<-as.data.frame(date.seq)
names(date.seq)<-'date' #make sure column name is 'date'

#Data frame 2 to have the counts and a Date variable
d3<-as.data.frame(d2)
d3$date<-as.Date(row.names(d3))

d4<- merge(d3, date.seq, all=T, by='date')
names(d4)<-c('date', 'a00_a09')
```

```{r}
plot(d4$date, d4$a00_a09, type='l', bty='l', ylab='Cases A00-A09')
```

## Now let's format the data for the surveillance package
Need to determine the year and week of the first data Can use week(date) and year(date) functions to do this

```{r}
lubridate::year(d4$date)
week(d4$date)
a00_DP <- create.disProg(
      week = 1:nrow(d4), #Index of observations
      observed = d4$a00_a09 ,
      state=matrix(0, nrow=nrow(d4), ncol=1),
      start = c(2000, 53))
```

Let's look for aberrations during the 2007 calendar years first, then see how it performs for 2008-2011 Need to  first determine indexes for thise year

Let's try historical limits with default threshold
```{r hist.limits, exercise=TRUE}
test.index<-which(lubridate::year(d4$date) %in% c(2007)) #returns vector of the eek numbers for 2008 and 2009

hl2<-algo.cdc(a00_DP,control = list(b = 5,  #N years in baseline
                                               m = 1, #N 4 week periods on either side
                                              alpha=0.05, #Signifiance level
                                               range=test.index)) #weeks tested

#This stuff just makes a pretty plot
col.alarms<-c(rep(1, times=(hl2$control$range[1]-1)) , (hl2$alarm+2))
cols<-c('gray', 'black', 'red')
all.agg<-rollapply(a00_DP$observed,4,sum, align='right', fill=NA)
plot(all.agg , pch=16 , bty='l', ylab='Cases', xlab='Week', col=cols[col.alarms])
points(c(rep(NA, times=(hl2$control$range[1]-1)) , hl2$upperbound), type='l')
title('Historical limits Cases vs threshold: alarms are RED')

```


Let's try Farrington method, with and without down-weighting past epidemics
```{r farrington1, echo=TRUE, exercise=TRUE}
test.index<-which(lubridate::year(d4$date) %in% c(2007)) #returns vector of the eek numbers for 2008 and 2009

mod1<-algo.farrington(a00_DP, #dataset to use
                control=list(range=test.index,
                b=5, #How many years of historical data to use
                w=3, #Number of weeks before and after current week to include in                                 model fitting
                reweight=TRUE, #Do you want to downweight past epidemics?
                plot=FALSE
                ))

col.alarms<-c(rep(1, times=(mod1$control$range[1])-1) , (mod1$alarm+2))
cols<-c('gray', 'black', 'red')
plot(d4$date,mod1$disProgObj$observed , pch=16 , bty='l', ylab='Cases', xlab='Week', col=cols[col.alarms])
points(d4$date[1:max(mod1$control$range)],c(rep(NA, times=(mod1$control$range[1]-1)) , mod1$upperbound), type='l')
title('Farrington. Cases vs threshold: alarms are RED')
```

```{r farrington1, echo=TRUE, exercise=TRUE}
test.index<-which(lubridate::year(d4$date) %in% c(2008:2011)) #returns vector of the eek numbers for 2008 and 2009

mod1<-algo.farrington(a00_DP, #dataset to use
                control=list(range=test.index,
                b=5, #How many years of historical data to use
                w=3, #Number of weeks before and after current week to include in                                 model fitting
                reweight=TRUE, #Do you want to downweight past epidemics?
                plot=FALSE,
                alpha=0.01
                ))

col.alarms<-c(rep(1, times=(mod1$control$range[1])-1) , (mod1$alarm+2))
cols<-c('gray', 'black', 'red')
plot(mod1$disProgObj$observed , pch=16 , bty='l', ylab='Cases', xlab='Week', col=cols[col.alarms])
points(c(rep(NA, times=(mod1$control$range[1]-1)) , mod1$upperbound), type='l')
title('Farrington. Cases vs threshold: alarms are RED')
```

There is a drop in cases apparent in the data in the last couple of years. How does this affect the algorithm?

